Batson Challenges Using Historical Selection Data

### 2021-07-20 analysis/paper_draft.Rmd
minor edits, spelling

Despite _Batson_ and its progeny, "prosecutors in criminal cases, and defense attorneys in civil case are more likely to use their challenges to dismiss African American jurors, whereas criminal defense attorneys and civil plaintiff's attorneys are more likely to exercise their strikes on white prospective jurors." [@Diamond2018]. This pattern persists, despite the fact that both sides' attorneys can often observe the race and gender of those in the venire who were and were not struck by peremptory challenge. The research literature suggests several explanations. These include a reluctance to raise a _Batson_ challenge on the premise that such a challenge raises with it the embarrassment, discomfort, and reputation costs with being challenged as racist [@OBrien2019, 29; @Offit2021]. Similarly, a trial judge deciding a _Batson_ challenge may similarly be reluctant to find against opposing counsel, particularly if that prosecutor repeatedly appears before that judge. Moreover, drawing inferences from historical data -- the use of perempetory challenges in _past_ cases -- is hard, because jury-selection records are not readily available and largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise require painstaking investigation [e.g., @Craft2018].

In theory, however, were an attorney's strike history in past cases easily available, such strike history can be compared to what we would expect to happen if, following the same procedure, venire members were struck at random [@Kadane2018; @Kadane2018a]. To do so, one would need to, for any peremptory-selection method, a corresponding likelihood function for how a prosecutor or defense attorney strikes prospective jurors of a protected class (e.g., racial-minorities). Yet, even with agreement on the appropriate statistical analysis, it is hard to complete the analysis in time to matter, unless "statisticans accompany attorneys in the jury selection process or a convenient programme or app is developed." [@Kadane2018, 12]

@@

Here, we discuss how we developed a software protoype for attorneys to assess whether one can credibly infer bias, for Batson purposes, based on an attorney's pattern of perempetory challenges in a particular case, either alone or in combination with historical data on that attorney's pattern of strikes in past cases. In doing so, we highlight two distinctive contributions.

@@

Under the "struck jury" method of peremptory challenges, the trial judge rules on all challenges for cause before the parties exercise any peremptories, so that the number of potential jurors who remain is the sum of number of seats on the jury (plus alternates, if any) and the number of peremptory challenges allotted to both sides. Accordingly, in a non-capital felony trial in federal district court using this method, the trial judge will, after excusing jurors for cause, make sure that there are 28 potential jurors remaining before the parties exercise their peremptory challenges (12 jurors plus 16 peremptories, assuming no alternate jurors). Then, the parties exercise their strikes on those 28 potential jurors. In a typical variant of the "struck jury" method, the parties strike in an alternating sequence, with the prosecution striking first. Where the number of strikes are not the same for the parties, an alternating sequence may result in one party exercising multiple strikes per turn. If a party decides not to use all its strikes, the jurors are seated based on the order in which they were selected or at random. Under "strike-and-replace" or "jury box" method, enough potential jurors are seated to fill the jury box. Some of them are excused for cause, and the parties decide whether or not to strike any of the remaining jurors with their peremptory challenges.  Once that is completed, new potential jurors take the seats of the excused or striken jurors, and the process repeats until all the jury seats are filled.  At any point, strikes can be used on anyone in the jury box, including those there previously but not struck.

Following @Kadane2018 and @Barrett2007, we model these peremptory-challenge processes as follows. For any given jury selection $i$, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors). In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$. If we define the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$ by adding one parameter $w$, we can measure the bias by different values of $w$. When $w=1$, it corresponds to the situation when there is no bias. When $w>1$, the probability of striking a cognizable class member is greater than non cognizable class, which means a discriminatory against cognizable class. On the contrary, when $w<1$, the probability of striking a cognizable class member is smaller than non cognizable class, which means a preference of cognizable class. However, under this set up, the weight of non cognizable class is the reciprocal of the weight of cognizable class, which is not a desirable character. Using $b = log(w)$ could avoid this issue. For any given value of the bias parameter $b$, the probability of strike od a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

When we use power prior incorporating the historical information, we can control the weight that we put on historical data by modifying the weight parameter $a$. In the simulation study, we also try different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show the influence of the weight parameter on modeling accuracy. We consider 735 scenarios in total, which is the combination of 7 different bias parameter of current data and of historical data, 3 different amount of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generate 1000 dataset and fit the model on the 1000 generated dataset. We record the mean value of the posterior mean and credible intervals of the 1000 model fits for each scenario. As for the credible interval, we tried three significance levels, 80\%, 90\% and 95\%. We also record the corresponding coverage rate. For instance, the 90\% coverage rate means the proportion of the 1000 model fits that the credible interval shows the true preference as we set. That is, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie on the right of 0 among the 1000 model fits. We show the coverage rate under different scenarios in the figures.

Figure \@ref(fig:b80) depicts the results of significance of estimated $b$ under 80\% credible interval. Figure \@ref(fig:b90) shows the results under 90\% credible interval and figure \@ref(fig:b95) shows the results under 95\% credible interval.

```{r b80,  echo=FALSE, fig.cap="80% significance of bias parameter", out.width = '100%'}
library(ggplot2)
# knitr::include_graphics("dino_plots.pdf")

# INSERT simulation and plot code
sims <- read.csv("sim_resultc.csv")
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L80 <0 & sims$U80 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L80 >0 & sims$U80 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L80 <0 & sims$U80 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

@@

# knitr::include_graphics("dino_plots.pdf")

@@

From the plot, our model can capture strong bias and bias, when the bias parameter being -3,-2,2,3. When the bias is moderate, that is when the bias parameter b being -1 and 1, the model can detect it when we have historical data with the same direction bias parameter. If the bias parameter of historical data is in an opposite direction, we might get a wrong estimate of current bias parameter. Thus if the historical data is not reliable, we should set a small weight parameter on them.

>>>>

Despite _Batson_ and its progeny, "prosecutors in criminal cases, and defense attorneys in civil case are more likely to use their challenges to dismiss African American jurors, whereas criminal defense attorneys and civil plaintiff's attorneys are more likely to exercise their strikes on white prospective jurors." [@Diamond2018]. This pattern persists, despite the fact that both sides' attorneys can often observe the race and gender of those in the venire who were and were not struck by peremptory challenge. The research literature suggests several explanations. These include a reluctance to raise a _Batson_ challenge on the premise that such a challenge raises with it the embarrassment, discomfort, and reputation costs with being challenged as racist [@OBrien2019, 29; @Offit2021]. Similarly, a trial judge deciding a _Batson_ challenge may similarly be reluctant to find against opposing counsel, particularly if that prosecutor repeatedly appears before that judge. Moreover, drawing inferences from historical data -- the use of peremptory challenges in _past_ cases -- is hard, because jury-selection records are not readily available and largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise require painstaking investigation [e.g., @Craft2018].

In theory, however, were an attorney's strike history in past cases easily available, such strike history can be compared to what we would expect to happen if, following the same procedure, venire members were struck at random [@Kadane2018; @Kadane2018a]. To do so, one would need to, for any peremptory-selection method, a corresponding likelihood function for how a prosecutor or defense attorney strikes prospective jurors of a protected class (e.g., racial-minorities). Yet, even with agreement on the appropriate statistical analysis, it is hard to complete the analysis in time to matter, unless "statisticians accompany attorneys in the jury selection process or a convenient programme or app is developed." [@Kadane2018, 12]

@@

Here, we discuss how we developed a software prototype for attorneys to assess whether one can credibly infer bias, for Batson purposes, based on an attorney's pattern of peremptory challenges in a particular case, either alone or in combination with historical data on that attorney's pattern of strikes in past cases. In doing so, we highlight two distinctive contributions.

@@

Under the "struck jury" method of peremptory challenges, the trial judge rules on all challenges for cause before the parties exercise any peremptories, so that the number of potential jurors who remain is the sum of number of seats on the jury (plus alternates, if any) and the number of peremptory challenges allotted to both sides. Accordingly, in a non-capital felony trial in federal district court using this method, the trial judge will, after excusing jurors for cause, make sure that there are 28 potential jurors remaining before the parties exercise their peremptory challenges (12 jurors plus 16 peremptories, assuming no alternate jurors). Then, the parties exercise their strikes on those 28 potential jurors.

In a typical variant of the "struck jury" method, the parties strike in an alternating sequence, with the prosecution striking first. Where the number of strikes are not the same for the parties, an alternating sequence may result in one party exercising multiple strikes per turn. If a party decides not to use all its strikes, the jurors are seated based on the order in which they were selected or at random. Under "strike-and-replace" or "jury box" method, enough potential jurors are seated to fill the jury box. Some of them are excused for cause, and the parties decide whether or not to strike any of the remaining jurors with their peremptory challenges.  Once that is completed, new potential jurors take the seats of the excused or stricken jurors, and the process repeats until all the jury seats are filled.  At any point, strikes can be used on anyone in the jury box, including those there previously but not struck.

Following @Kadane2018 and @Barrett2007, we model these peremptory-challenge processes as follows. For any given jury selection $i$, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors).

In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike.

If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$. If we define the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$ by adding one parameter $w$, we can measure the bias by different values of $w$. When $w=1$, it corresponds to the situation when there is no bias. When $w>1$, the probability of striking a cognizable class member is greater than non cognizable class, which means a discriminatory against cognizable class. On the contrary, when $w<1$, the probability of striking a cognizable class member is smaller than non cognizable class, which means a preference of cognizable class.

However, under this set up, the weight of non cognizable class is the reciprocal of the weight of cognizable class, which is not a desirable character. Using $b = log(w)$ could avoid this issue. For any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

When we use power prior incorporating the historical information, we can control the weight that we put on historical data by modifying the weight parameter $a$. In the simulation study, we also try different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show the influence of the weight parameter on modeling accuracy. We consider 735 scenarios in total, which is the combination of 7 different bias parameter of current data and of historical data, 3 different amount of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generate 1000 datasets and fit the model on the 1000 generated dataset. We record the mean value of the posterior mean and credible intervals of the 1000 model fits for each scenario.

For the credible interval, we set three significance levels (80\%, 90\% and 95\%) and recorded the corresponding coverage rate. For instance, the 90\% coverage rate means the proportion of the 1000 model fits that the credible interval shows the true preference as we set. That is, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie on the right of 0 among the 1000 model fits.

The following figures depict the coverage rate under different scenarios. Figure \@ref(fig:b80) depicts the results of significance of estimated $b$ under 80\% credible interval. Figure \@ref(fig:b90) shows the results under 90\% credible interval and figure \@ref(fig:b95) shows the results under 95\% credible interval.

```{r b80,  echo=FALSE, fig.cap="80% significance of bias parameter", out.width = '100%'}

library(ggplot2)

# INSERT simulation and plot code

load_path <- here::here("sim_resultc.csv")

sims <- readr::read_csv(load_path)
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L80 <0 & sims$U80 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L80 >0 & sims$U80 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L80 <0 & sims$U80 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

@@

@@

From the plot, our model can capture strong bias and bias, when the bias parameter being -3,-2,2,3. When bias is moderate ($b = \{1, -1\}$), the model can detect it when we have historical data with the same direction bias parameter. If the bias parameter of historical data is in an opposite direction, we might get a wrong estimate of current bias parameter.

Thus if the historical data is not reliable, we should set a small weight parameter on them. For example, [insert]
